\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A Radial Time Projection Chamber for $\alpha$ detection in CLAS at JLab}

### 2017-07-05 eg6-letters/NIM/EG6_RTPC_NIM.tex
New wave of Stepan's comments

The original design of the RTPC was developed for the BoNuS 
experiment at Jefferson Lab which took data in CLAS in 2005~\cite{BONUS-NIM}. An enhanced design, 
used in the 2009 DVCS experiment, is described in section \ref{sec_design} 
of this paper. Significant improvements were made to the RTPC mechanical 
structure and fabrication technique that both increased the acceptance and 
reduced the amount of material in the path of the outgoing particles. The 
data acquisition system described in section \ref{sec_readout} was reconfigured 
to increase the event readout rate. The tracking algorithm is discussed in section
\ref{sec_tracking}. The calibration methods 
are discussed in section \ref{sec_calib}. Finally, the overall performance
of the RTPC is described in section \ref{sec_perfor}.

@@

We detail here the different regions shown in Fig.~\ref{fig:RTPC_1_4} starting 
from the beam line towards larger radius:\\
\begin{itemize}
   \item The 6 atm helium-4 target extends along the beam line axis; it 
      is a 284~mm long, 6~mm diameter straw with a 27-$\mu$m Kapton wall.
   \item The first gas gap covers a radial range from 3 mm to 20 mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o{}ller electrons scattered by the beam. This region is surrounded by 
      a 4 $\mu$m thick window made of grounded aluminized Mylar.
   \item The second gap region extends between 20~mm and 30~mm and is filled with the 
      gas mixture of 80$\%$ neon (Ne) and 20$\%$ dimethyl ether (DME). This region 
      is surrounded by a 4~$\mu$m thick window made of aluminized Mylar set at $- 4260$~V 
      to serve as the cathode.
   \item The drift region is filled with the same Ne-DME gas mixture and extends 
      from the cathode to the first GEM, 60 mm away from the beam axis. The average 
      electric field in this region is perpendicular to the beam and about 550~V/cm.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66 mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower voltage to
      obtain a strong ($\sim$1600 V/cm) electric field between the GEM foils. A 
      275~V bias is applied across each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charge
     from the GEMs. Pre-amplifiers are plugged directly on its outer side and
     transmit the signal to the data acquisition electronics.
\end{itemize}

@@

In this section we will detail the methods used to calibrate the drift speed,
drift paths and gains of the detector. Drift speed and paths were initially
calculated using the MAGBOLTZ~\cite{MAGBOLTZ} program, then refined using
data to account for variations of the run conditions. We always assume 
cylindrical symmetry in the chamber for the calibration, such that none of
the parameters depend on the azimuthal angle $\phi$. The initial MAGBOLTZ
calibration was improved through several iterations of the
process described below, each iteration increasing the number of events 
properly reconstructed in the RTPC. The figures presented in this section
are the ones obtained while performing the last iteration of this
calibration process.

@@

After the ionization, the released electrons 
drift to the cylindrical detection plane under the effect of the electric field. The 
electrons released close to the cathode take the most time to reach the readout 
pads, but cylindrical symmetry insures that they always travel the same 
distance. We illustrate this in Fig.~\ref{fig:RTPC_signals}, where we depict a 
typical $^{4}$He track. By measuring the maximum time ($T_{Max}$), we can infer the drift 
speed of the electrons in the RTPC.\\

@@

Due to the non perfect experimental conditions, in particular possible contaminations 
of our gas mixture, the drift speed changed during the three months long experimental 
run\footnote{We had to increase the gas flow during the experiment due to a small leak 
in the RTPC, which concurred to the large shift of drift time observed around run number 
61600. While our gas system was kept slightly over 
atmospheric pressure to avoid contamination from air or other external gases, it 
is likely that this leak was the source of modification of the gas mixture.}.  
Figure~\ref{fig:Drift_run_number_1} shows the $T_{Max/2}$ values for 
individual runs (approximately 2 hours long). We observe a significant 
variation of the drift speed over time that need to be accounted for 
in track reconstruction.

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. We calculate them with
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used such calculation as a 
first calibration, but, 
as can be seen with the drift speed, experimental conditions were unstable in the 
chamber. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometry accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.204 and 1.269 GeV) to enhance the cross section of the elastic 
scattering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematics allows to calculate directly the kinematics of the Helium nucleus. 
By comparing the calculated momentum and angle of the recoil alpha particle to the 
measurement in the RTPC, we tuned the drift paths independently of our 
knowledge of the chamber's conditions.

@@

To calibrate the gains, we compare the experimental ADCs to the energy deposited 
for each pad individually in GEANT4 by similar simulated tracks (using the 
same elastic events as for the drift 
paths calibration). This requires a very good GEANT4 simulation 
including drift paths, but also the spread of the charges along the path
before reaching the pad, so that the simulated hits match the experimental 
ones. Moreover, the simulation has to match the DAQ features 
that can lead to cutting out hits. After setting the simulation properly, we 
compared simulation to experiment on an event by event basis as shown in 
Fig.~\ref{fig:EVENT_adc_tdc}. The gain for each pad is calculated 
as the ratio of the measured ADCs to the simulated deposited energy.  
Then, these gains are refined using correction factors obtained from a 
sample of good tracks. For each track, we calculate the corrected 
energy deposit on a pad and compare it to the average
deposit recorded by the other pads. This last step was implemented to 
avoid any bias from this simulation based calibration method.
Final results are shown in Fig.~\ref{fig:dedx_p_exp_2nd}, where energy 
loss of particles is plotted against momentum over charge ration. One can 
clearly see the band for $^4$He in its expected position.

@@

The first is an oscillatory noise located early in the readout time window, 
shown in the top panel of Fig.~\ref{fig:noise} for a particularly noisy 
channel. Its amplitude is not dissimilar to those of real tracks. About 
18\% of the readout channels exhibit large contributions from this noise 
characteristic. Due to its unique time-energy correlation for the given 
channels, the noise could be removed on an event by event and channel by 
channel basis without significant loss of good signals. The result of the
procedure is illustrated in the bottom panel of Fig.~\ref{fig:noise}.

The second noise signature was a coherent noise affecting about 25\% of the 
amplifiers boards, when simultaneous hits in most of the 16 channels of the
board were recorded. An event-based technique to identify and remove this noise was developed based 
on counting simultaneous hits in preamp group, and, if sufficiently large, 
perform a dynamic pedestal subtraction based on the average ADC of the 
channels' neighbors within the preamp group.

The sources of these effects were not determined, but rejection techniques 
allowed to reconstruct 10\% more good tracks, with no significant loss, and 
recover 70 channels that were previously ignored due to excessive noise levels.

@@

\subsection{Track Fitting}\label{sec_rec}
We start with reconstructing the spacial origin of the hits using the extracted drift speed and drift path 
parameters. For each registered hit, we calculate a position of emission from 
the recorded time and the position of the recording pad. The third step 
is to create chains of hits. The maximum distance between two close adjacent 
hits has to be less than 10.5 mm to chain them, which roughly corresponds to 
neighbors and next to neighbors. Then, we fit with a helix the chains that have a minimum of 
10 hits. We make the fit in two iterations, first with all the hits of the chain 
and the beam line, then, in the second iteration, 
the hits that are 5 mm or farther from the first fit are excluded.

@@

\subsection{Efficiency}
We measured the efficiency of the RTPC
using elastic scattering on $^4$He by comparing the inclusive yield, based
only on electron detection, to the exclusive elastic yield, where the Helium
recoil is also detected (see Fig.~\ref{fig:w}). We present in Fig.~\ref{fig:rtpc_eff}
the results for the two halves of the detector. We observe that the left and 
the right modules have similar efficiencies except near the upstream target 
window. This difference is due to the large number of dead channels in this 
area of the detector, which are concentrated on the left.

>>>>

The original design of the RTPC was developed for the BoNuS 
experiment at Jefferson Lab which took data with CLAS in 2005~\cite{BONUS-NIM}. An enhanced design, 
used in the 2009 DVCS experiment, is described in section \ref{sec_design} 
of this paper. Significant improvements were made to the RTPC mechanical 
structure and fabrication technique that both increased the acceptance and 
reduced the amount of material in the path of the outgoing particles. The 
data acquisition system described in section \ref{sec_readout} was reconfigured 
to increase the event readout rate. The calibration methods 
are discussed in section \ref{sec_calib}. The tracking algorithm is discussed in section
\ref{sec_tracking}. Finally, the overall performance
of the RTPC is described in section \ref{sec_perfor}.

@@

We detail here the different regions shown in Fig.~\ref{fig:RTPC_1_4} starting 
from the beam line towards larger radius:\\
\begin{itemize}
   \item The 6 atm helium-4 target extends along the beam axis; it 
      is a 284~mm long, 6~mm diameter straw with a 27-$\mu$m Kapton wall.
   \item The first gas gap covers the radial range from 3 mm to 20 mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o{}ller electrons scattered by the beam. This region is surrounded by 
      a 4 $\mu$m thick window made of grounded aluminized Mylar.
   \item The second gap region extends between 20~mm and 30~mm and is filled with the 
      gas mixture of 80$\%$ neon (Ne) and 20$\%$ dimethyl ether (DME). This region 
      is surrounded by a 4~$\mu$m thick window made of aluminized Mylar set at $- 4260$~V 
      to serve as the cathode.
   \item The drift region is filled with the same Ne-DME gas mixture and extends 
      from the cathode to the first GEM, 60 mm away from the beam axis. The average 
      electric field in this region is perpendicular to the beam and about 550~V/cm.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66 mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower voltage to
      obtain a strong ($\sim$1600 V/cm) electric field between the GEM foils. A 
      275~V bias is applied across each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charge
     from the GEMs. Pre-amplifiers are plugged directly on its outer side and
     transmit the signal to the data acquisition electronics.
\end{itemize}

@@

In this section we will detail the methods used to calibrate the drift speed,
drift paths and gains of the detector. Drift speed and paths were initially
calculated using the MAGBOLTZ~\cite{MAGBOLTZ} program, then refined using
data to account for variations of the run conditions. We always assume 
cylindrical symmetry in the chamber for the calibration, such that none of
the parameters depend on the azimuthal angle $\phi$. The initial MAGBOLTZ
calibration was improved through several iterations of the
process described below, with each time an increasing number of events 
properly reconstructed in the RTPC. The figures presented in this section
are the ones obtained while performing the last iteration of this
calibration process.

@@

After the ionization, the released electrons 
drift to the cylindrical detection plane under the effect of the electric field. The 
electrons released close to the cathode take the most time to reach the readout 
pads, the cylindrical symmetry insures that the maximum distance is the same for all 
tracks. We illustrate this in Fig.~\ref{fig:RTPC_signals}, where we depict a 
typical $^{4}$He track. By measuring the maximum time ($T_{Max}$), we can infer the drift 
speed of the electrons in the RTPC.\\

@@

Due to the non perfect experimental conditions, in particular possible contamination 
of our gas mixture, the drift speed changed during the three months long experimental 
run\footnote{We had to increase the gas flow during the experiment due to a small leak 
in the RTPC, which concurred to the large shift of drift time observed around run number 
61600. While our gas system was kept slightly over 
atmospheric pressure to avoid contamination from air or other external gases, it 
is likely that this leak was the source of modification of the gas mixture.}.  
Figure~\ref{fig:Drift_run_number_1} shows the $T_{Max/2}$ values for 
individual runs (approximately 2 hours long). We observe significant change in 
the drift speed before and after run 61600, while variations within these periods 
are ~2\%. These variations of the drift speed were accounted in the track reconstruction

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. We calculate them with
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used such calculation as a 
first calibration, but, 
as can be seen with the drift speed, conditions in the chamber were changing 
over time. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometry accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.204 and 1.269 GeV) to enhance the cross section of the elastic 
scattering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematics allows to calculate directly the kinematics of the Helium nucleus. 
By comparing the calculated momentum and angle of the recoil alpha particle to the 
measurement in the RTPC, we tuned the drift paths independently of our 
knowledge of the chamber's conditions.

@@

To calibrate the gains, we compare the experimental ADCs to the energy deposited 
for each pad individually in GEANT4 by similar simulated tracks (using the 
same elastic events as for the drift 
paths calibration). This requires a very good GEANT4 simulation 
including drift paths, but also the spread of the charge along the path
before reaching the pad, so that the simulated hits match the experimental 
ones. Moreover, the simulation has to match the DAQ features 
that can lead to cutting out hits. After setting the simulation properly, we 
compared simulation to experiment on an event by event basis as shown in 
Fig.~\ref{fig:EVENT_adc_tdc}. The gain for each pad is calculated 
as the ratio of the measured ADCs to the simulated deposited energy.  
Then, these gains are refined using correction factors obtained from a 
sample of good tracks. For each track, we calculate the corrected 
energy deposit on a pad and compare it to the average
deposit recorded by the other pads. This last step was implemented to 
avoid any bias from this simulation based calibration method.
Final results are shown in Fig.~\ref{fig:dedx_p_exp_2nd}, where energy 
loss of particles is plotted against momentum over charge ratio. One can 
clearly see the band for $^4$He in its expected position.

@@

The first is an oscillatory noise located early in the readout time window, 
shown in the top panel of Fig.~\ref{fig:noise} for a particularly noisy 
channel. Its amplitude is similar to those of real tracks. About 
18\% of the readout channels exhibit large contributions from this noise 
characteristic. Due to its unique time-energy correlation for the given 
channels, the noise could be removed on an event by event and channel by 
channel basis without significant loss of good signals. The result of the
procedure is illustrated in the bottom panel of Fig.~\ref{fig:noise}.

The second noise signature was a coherent noise affecting about 25\% of the 
amplifiers boards, when simultaneous hits in most of the 16 channels of the
board were recorded. An event-based technique to identify and remove this noise was developed based 
on counting simultaneous hits in each pre-amplifier group, and, if sufficiently large, 
perform a dynamic pedestal subtraction based on the average ADC of neighboring 
channels within this group.

The sources of these effects were not determined, but rejection techniques 
allowed to reconstruct 10\% more good tracks and 
recover 70 channels that were previously ignored due to excessive noise levels.

@@

\subsection{Track Fitting}\label{sec_rec}
We start with reconstructing the spacial origin of the hits using the extracted drift speed and drift path 
parameters. For each registered hit, we calculate a position of emission from 
the recorded time and the position of the recording pad. The third step 
is to create chains of hits. The maximum distance between two close adjacent 
hits has to be less than 10.5 mm to chain them, which roughly corresponds to 
neighbors and next to neighbors. Then, we fit the chains with a helix if they have a minimum of 
10 hits. We then eliminate from the chain the hits that are 5 mm or farther from the fit
as they are not likely part of the same track.
This new reduced chain is used for a second and final helix fit.

@@

\subsection{Efficiency}
We measured the efficiency of the RTPC
using elastic scattering on $^4$He by comparing the inclusive yield, based
only on electron detection, to the exclusive elastic yield, where the Helium
recoil is also detected (see Fig.~\ref{fig:w}). We present in Fig.~\ref{fig:rtpc_eff}
the results for the two halves of the detector. We observe that the left and 
the right modules have similar efficiencies except near the upstream target 
window. This difference is due to the large number of dead channels concentrated
in this part of the left half of the detector
